---
title: "Poisson Regression and Maximum Likelihood Estimation"
subtitle: "A Case Study of Blueprinty's Software and Patent Awards"
author: "Kristin Yang"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

.libPaths(c("~/R/library", .libPaths()))

library(ggplot2)
library(dplyr)
library(knitr)
library(scales)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(color = "#4B5563"),
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold")
    )
)

blueprinty <- read.csv("blueprinty.csv") |>
  mutate(
    customer_status = if_else(iscustomer == 1, "Blueprinty customer", "Non-customer"),
    customer_status = factor(customer_status, levels = c("Non-customer", "Blueprinty customer")),
    region = factor(region)
  )

Y <- blueprinty$patents
patent_means <- blueprinty |>
  group_by(customer_status) |>
  summarize(
    firms = n(),
    mean_patents = mean(patents),
    median_patents = median(patents),
    .groups = "drop"
  )
```

## Introduction

Blueprinty is a small software company that helps engineering firms create blueprint materials for patent applications. Its marketing team would like to make a strong claim: firms that use Blueprinty's software are more successful at getting patents awarded.

That claim sounds simple, but the evidence is not as clean as we would like. The ideal dataset would follow the same firms before and after they adopted Blueprinty's product, or randomly assign firms to use the software. Instead, Blueprinty has collected an observational dataset on `r nrow(blueprinty)` mature engineering firms. For each firm, we observe the number of patents awarded over the last five years, its region, its age since incorporation, and whether it is a Blueprinty customer.

The key challenge is selection. Blueprinty's customers were not selected at random. If customers are older, located in regions with more patent activity, or otherwise systematically different from non-customers, then a raw difference in patent counts could reflect those differences rather than the software itself. The goal of this post is to use maximum likelihood estimation (MLE) and Poisson regression to make the comparison more disciplined: first by modeling patent counts as counts, and then by adjusting for observable firm characteristics.

## Exploring the Data

The outcome is the number of patents awarded to each firm over a five-year window. Customers do appear to patent more in the raw data: Blueprinty customers average `r round(patent_means$mean_patents[patent_means$customer_status == "Blueprinty customer"], 2)` patents, compared with `r round(patent_means$mean_patents[patent_means$customer_status == "Non-customer"], 2)` among non-customers.

```{r}
#| label: patent-summary-table
#| echo: false

patent_means |>
  mutate(mean_patents = round(mean_patents, 2)) |>
  rename(
    `Firm type` = customer_status,
    `Number of firms` = firms,
    `Mean patents` = mean_patents,
    `Median patents` = median_patents
  ) |>
  kable(
    caption = "Patent awards over the last five years by customer status",
    align = "lrrr",
    format = "html",
    table.attr = 'class="table table-striped table-hover table-sm"'
  )
```

```{r}
#| label: patent-histograms
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| fig-cap: "Patent count distributions by Blueprinty customer status"

ggplot(blueprinty, aes(x = patents, fill = customer_status)) +
  geom_histogram(binwidth = 1, boundary = -0.5, color = "white", alpha = 0.9) +
  geom_vline(
    data = patent_means,
    aes(xintercept = mean_patents),
    color = "#111827",
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  facet_wrap(~customer_status, nrow = 1) +
  scale_x_continuous(breaks = seq(0, max(blueprinty$patents), by = 2)) +
  scale_fill_manual(values = c("#2F6F73", "#C8553D")) +
  labs(
    title = "Blueprinty customers have a visibly higher patent distribution",
    subtitle = "Dashed vertical lines mark group means",
    x = "Patents awarded over the last five years",
    y = "Number of firms",
    fill = NULL
  ) +
  theme(legend.position = "none")
```

The customer histogram is shifted slightly to the right. That is encouraging for Blueprinty's marketing claim, but it is not yet convincing. The next question is whether customers and non-customers look comparable on other characteristics that may also predict patenting.

One major imbalance is region. More than two-thirds of Blueprinty's customers are in the Northeast, while non-customers are spread more evenly across regions. If Northeast firms are different in their patenting behavior for reasons unrelated to Blueprinty, failing to control for region would blur the interpretation of the customer effect.

```{r}
#| label: region-balance
#| echo: false
#| fig-width: 8.5
#| fig-height: 5
#| fig-cap: "Regional composition by customer status"

region_balance <- blueprinty |>
  count(customer_status, region) |>
  group_by(customer_status) |>
  mutate(share = n / sum(n)) |>
  ungroup()

ggplot(region_balance, aes(x = customer_status, y = share, fill = region)) +
  geom_col(width = 0.68, color = "white") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Blueprinty customers are disproportionately located in the Northeast",
    x = NULL,
    y = "Share of firms",
    fill = "Region"
  ) +
  theme(legend.position = "right")
```

```{r}
#| label: region-table
#| echo: false

region_balance |>
  mutate(`Share within firm type` = percent(share, accuracy = 0.1)) |>
  select(
    `Firm type` = customer_status,
    Region = region,
    Firms = n,
    `Share within firm type`
  ) |>
  kable(
    caption = "Regional composition of customers and non-customers",
    align = "llrr",
    format = "html",
    table.attr = 'class="table table-striped table-hover table-sm"'
  )
```

Firm age is less dramatically imbalanced, but it still matters. Customers are slightly older on average (`r round(mean(blueprinty$age[blueprinty$iscustomer == 1]), 1)` years vs. `r round(mean(blueprinty$age[blueprinty$iscustomer == 0]), 1)` years). Since older firms may have more developed R&D operations, existing legal processes, or deeper patent portfolios, age belongs in the model too.

```{r}
#| label: age-histograms
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| fig-cap: "Firm age distributions by Blueprinty customer status"

age_means <- blueprinty |>
  group_by(customer_status) |>
  summarize(mean_age = mean(age), .groups = "drop")

ggplot(blueprinty, aes(x = age, fill = customer_status)) +
  geom_histogram(bins = 24, color = "white", alpha = 0.9) +
  geom_vline(
    data = age_means,
    aes(xintercept = mean_age),
    color = "#111827",
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  facet_wrap(~customer_status, nrow = 1) +
  scale_fill_manual(values = c("#2F6F73", "#C8553D")) +
  labs(
    title = "Customers and non-customers have similar, but not identical, age profiles",
    subtitle = "Dashed vertical lines mark group means",
    x = "Firm age in years",
    y = "Number of firms",
    fill = NULL
  ) +
  theme(legend.position = "none")
```

The exploratory picture is therefore mixed: customers have more patents, but customers are also concentrated in one region and differ slightly in age. This is exactly the kind of setting where regression is useful. It will not magically create a randomized experiment, but it can compare customers and non-customers after holding observed differences constant.

## A Simple Poisson Model via Maximum Likelihood

Patent awards are count data. Counts are non-negative integers, so a Normal model is an awkward starting point: it can assign probability to impossible negative or fractional outcomes. A Poisson distribution is a natural first model for this type of outcome because it is built for non-negative integer counts.

Start with the simplest possible model:

$$
Y_i \sim \text{Poisson}(\lambda)
$$

where every firm is assumed to share the same patent-award rate $\lambda$. For one observation,

$$
f(Y_i|\lambda) = \frac{e^{-\lambda}\lambda^{Y_i}}{Y_i!}.
$$

Assuming the observations are independent and identically distributed, the joint likelihood is the product of the individual probabilities:

$$
L(\lambda|Y_1,\ldots,Y_n)
= \prod_{i=1}^{n} \frac{e^{-\lambda}\lambda^{Y_i}}{Y_i!}.
$$

We can simplify that product:

$$
L(\lambda|Y)
= \frac{e^{-n\lambda}\lambda^{\sum_i Y_i}}{\prod_i Y_i!}.
$$

The log-likelihood is easier to work with:

$$
\ell(\lambda|Y)
= -n\lambda + \left(\sum_i Y_i\right)\log(\lambda) - \sum_i \log(Y_i!).
$$

The last term does not involve $\lambda$, but it is still part of the log-likelihood value. In code, it is safest to let R's built-in Poisson density handle the factorial term on the log scale.

```{r}
#| label: simple-poisson-loglikelihood
#| echo: true
#| code-fold: false

poisson_loglikelihood <- function(lambda, Y) {
  if (lambda <= 0) {
    return(-Inf)
  }
  sum(dpois(Y, lambda = lambda, log = TRUE))
}

lambda_grid <- seq(0.25, 8, length.out = 400)
loglik_grid <- data.frame(
  lambda = lambda_grid,
  log_likelihood = sapply(lambda_grid, poisson_loglikelihood, Y = Y)
)

simple_mle <- optim(
  par = 1,
  fn = function(lambda) -poisson_loglikelihood(lambda, Y),
  method = "L-BFGS-B",
  lower = 1e-8
)

lambda_hat_optim <- simple_mle$par
lambda_hat_mean <- mean(Y)
```

```{r}
#| label: simple-poisson-plot
#| echo: true
#| code-fold: false
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Log-likelihood for the single-parameter Poisson model"

ggplot(loglik_grid, aes(x = lambda, y = log_likelihood)) +
  geom_line(color = "#2F6F73", linewidth = 1.1) +
  geom_vline(xintercept = lambda_hat_optim, color = "#C8553D", linewidth = 0.9) +
  annotate(
    "label",
    x = lambda_hat_optim + 1.1,
    y = max(loglik_grid$log_likelihood) - 40,
    label = paste0("MLE = ", round(lambda_hat_optim, 3)),
    fill = "white",
    color = "#111827",
    label.size = 0.2
  ) +
  labs(
    title = "The MLE is the peak of the log-likelihood curve",
    x = expression(lambda),
    y = "Log-likelihood"
  )
```

Visually, the MLE is the value of $\lambda$ where the curve reaches its highest point. Here, that point is almost exactly the sample mean.

```{r}
#| label: simple-mle-table
#| echo: false

data.frame(
  Method = c("Numerical MLE from optim()", "Sample mean"),
  Estimate = c(lambda_hat_optim, lambda_hat_mean)
) |>
  mutate(Estimate = round(Estimate, 6)) |>
  kable(
    caption = "The Poisson MLE for a common patent rate",
    align = "lr",
    format = "html",
    table.attr = 'class="table table-striped table-hover table-sm"'
  )
```

The equality is not a coincidence. Taking the derivative of the log-likelihood gives

$$
\frac{\partial \ell}{\partial \lambda}
= -n + \frac{\sum_i Y_i}{\lambda}.
$$

Setting this equal to zero:

$$
-n + \frac{\sum_i Y_i}{\lambda} = 0
\qquad \Rightarrow \qquad
\hat{\lambda}_{MLE} = \frac{1}{n}\sum_i Y_i = \bar{Y}.
$$

This result feels right because the mean of a Poisson distribution is $\lambda$. If all firms share one patent rate, the best estimate of that rate is simply the average patent count.

## Poisson Regression

The one-parameter model is too simple for Blueprinty's business question because it assumes every firm has the same expected number of patents. A Poisson regression lets each firm's rate depend on its characteristics:

$$
Y_i \sim \text{Poisson}(\lambda_i),
\qquad
\lambda_i = \exp(X_i'\beta).
$$

The exponential link is important. The linear index $X_i'\beta$ can be any real number, but a Poisson rate must be positive. Exponentiating maps the model's prediction back onto the positive real line.

For the regression model, the log-likelihood becomes

$$
\ell(\beta|Y,X)
= \sum_{i=1}^{n}
\left[
Y_i X_i'\beta - \exp(X_i'\beta) - \log(Y_i!)
\right].
$$

The covariate matrix below includes an intercept, firm age, firm age squared, region indicators, and the customer indicator. Age squared allows the age-patenting relationship to bend instead of forcing it to be perfectly linear. R automatically drops one region category, so the region coefficients are interpreted relative to the omitted baseline, which is Midwest in this dataset.

```{r}
#| label: poisson-regression-mle
#| echo: true
#| code-fold: false

poisson_regression_loglikelihood <- function(beta, Y, X) {
  eta <- as.vector(X %*% beta)
  sum(Y * eta - exp(eta) - lgamma(Y + 1))
}

poisson_regression_score <- function(beta, Y, X) {
  eta <- as.vector(X %*% beta)
  lambda <- exp(eta)
  as.vector(crossprod(X, Y - lambda))
}

X <- model.matrix(~ age + I(age^2) + region + iscustomer, data = blueprinty)

start_beta <- c(log(mean(Y)), rep(0, ncol(X) - 1))
names(start_beta) <- colnames(X)

ndeps <- rep(1e-3, ncol(X))
ndeps[colnames(X) == "age"] <- 1e-4
ndeps[colnames(X) == "I(age^2)"] <- 1e-6

poisson_mle <- optim(
  par = start_beta,
  fn = function(beta) -poisson_regression_loglikelihood(beta, Y, X),
  gr = function(beta) -poisson_regression_score(beta, Y, X),
  method = "BFGS",
  hessian = TRUE,
  control = list(maxit = 10000, reltol = 1e-12, ndeps = ndeps)
)

beta_hat <- poisson_mle$par
se_hat <- sqrt(diag(solve(poisson_mle$hessian)))
```

Because `optim()` minimizes functions, I minimized the negative log-likelihood. The Hessian returned by `optim()` is therefore the Hessian of the negative log-likelihood, also called the observed information matrix. The coefficient standard errors are the square roots of the diagonal of its inverse.

As a check, I fit the same model using R's built-in `glm()` function.

```{r}
#| label: glm-check
#| echo: true
#| code-fold: false

poisson_glm <- glm(
  patents ~ age + I(age^2) + region + iscustomer,
  family = poisson(link = "log"),
  data = blueprinty
)
```

```{r}
#| label: regression-table
#| echo: false

term_labels <- c(
  "(Intercept)" = "Intercept",
  "age" = "Age",
  "I(age^2)" = "Age squared",
  "regionNortheast" = "Northeast vs. Midwest",
  "regionNorthwest" = "Northwest vs. Midwest",
  "regionSouth" = "South vs. Midwest",
  "regionSouthwest" = "Southwest vs. Midwest",
  "iscustomer" = "Blueprinty customer"
)

glm_summary <- summary(poisson_glm)$coefficients

regression_table <- data.frame(
  Term = term_labels[colnames(X)],
  `MLE estimate` = beta_hat,
  `MLE SE` = se_hat,
  `glm() estimate` = coef(poisson_glm),
  `glm() SE` = glm_summary[, "Std. Error"],
  check.names = FALSE
)

regression_table |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  kable(
    caption = "Poisson regression estimates from manual MLE and glm()",
    align = "lrrrr",
    format = "html",
    table.attr = 'class="table table-striped table-hover table-sm"'
  )
```

The manual MLE and `glm()` estimates match closely, which is the main sanity check. The coefficient on `iscustomer` is positive: holding age and region fixed, Blueprinty customers have a higher expected patent count. Since Poisson regression is log-linear, the coefficient is best read multiplicatively. A coefficient of `r round(beta_hat["iscustomer"], 3)` corresponds to a rate ratio of $\exp(`r round(beta_hat["iscustomer"], 3)`)$, or about `r round((exp(beta_hat["iscustomer"]) - 1) * 100, 1)`% higher expected patent awards for customers with the same age and region.

```{r}
#| label: rate-ratio-table
#| echo: false

rate_ratio_table <- data.frame(
  Term = term_labels[colnames(X)][-1],
  `Rate ratio` = exp(beta_hat[-1]),
  `Percent change` = (exp(beta_hat[-1]) - 1) * 100,
  `Lower 95% CI` = exp(beta_hat[-1] - 1.96 * se_hat[-1]),
  `Upper 95% CI` = exp(beta_hat[-1] + 1.96 * se_hat[-1]),
  check.names = FALSE
)

rate_ratio_table |>
  mutate(
    `Rate ratio` = round(`Rate ratio`, 3),
    `Percent change` = paste0(round(`Percent change`, 1), "%"),
    `Lower 95% CI` = round(`Lower 95% CI`, 3),
    `Upper 95% CI` = round(`Upper 95% CI`, 3)
  ) |>
  kable(
    caption = "Multiplicative interpretation of the Poisson coefficients",
    align = "lrrrr",
    format = "html",
    table.attr = 'class="table table-striped table-hover table-sm"'
  )
```

The age terms are also sensible. The positive age coefficient and negative age-squared coefficient imply that expected patenting rises with age at first, then flattens and eventually declines. The implied peak is around `r round(-beta_hat["age"] / (2 * beta_hat["I(age^2)"]), 1)` years, which suggests that the most patent-active firms in this sample are neither very young nor very old. The region estimates are small after controlling for age and customer status, which means the most important region fact for this analysis is not that regions have enormous conditional effects, but that customers and non-customers are distributed very differently across them.

## The Effect of Blueprinty's Software

The regression coefficient is useful, but it is still in log-rate units. For a marketing question, "how many more patents?" is easier to understand than "how much larger is the log expected count?" To translate the model into patent counts, I construct two counterfactual datasets:

1. $X_0$: every firm is treated as a non-customer.
2. $X_1$: every firm is treated as a Blueprinty customer.

Age and region stay exactly as observed for each firm. Only the customer indicator changes.

```{r}
#| label: counterfactual-effect
#| echo: true
#| code-fold: false

X0 <- X
X1 <- X

X0[, "iscustomer"] <- 0
X1[, "iscustomer"] <- 1

yhat0 <- as.vector(exp(X0 %*% beta_hat))
yhat1 <- as.vector(exp(X1 %*% beta_hat))

average_customer_effect <- mean(yhat1 - yhat0)
counterfactual_means <- c(
  `If no firms used Blueprinty` = mean(yhat0),
  `If all firms used Blueprinty` = mean(yhat1),
  `Average difference` = average_customer_effect
)
```

```{r}
#| label: counterfactual-table
#| echo: false

data.frame(
  Counterfactual = names(counterfactual_means),
  `Predicted patents per firm` = as.numeric(counterfactual_means),
  check.names = FALSE
) |>
  mutate(`Predicted patents per firm` = round(`Predicted patents per firm`, 3)) |>
  kable(
    caption = "Counterfactual predicted patent counts over five years",
    align = "lr",
    format = "html",
    table.attr = 'class="table table-striped table-hover table-sm"'
  )
```

The model estimates that being a Blueprinty customer is associated with about `r round(average_customer_effect, 2)` additional patents per firm over five years, holding age and region fixed. That is larger than a trivial difference: for a firm deciding whether software improves the patent-application process, an expected gain of roughly four-fifths of a patent over five years could matter.

The careful wording is "associated with," not "caused by." This analysis supports Blueprinty's marketing story in the sense that customers have higher patent counts even after adjusting for age and region. But the data are observational. We still cannot rule out unobserved differences: Blueprinty customers may have stronger legal teams, larger R&D budgets, better patent strategy, or more aggressive management. A randomized experiment or credible before-and-after design would be needed to make a stronger causal claim. As evidence for a marketing argument, though, the Poisson regression gives Blueprinty a much more credible story than the raw comparison alone.
